---
title: "FOXP2 promoter reanalysis (Mackenzie 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FOXP2 promoter reanalysis (Mackenzie 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

# Overview

This vignette walks through the FOXP2 promoter reanalysis from the
GLproxScape paper, using the dataset of Mackenzie et al. (2026, *Mol Cell
Proteomics*): three sgRNAs tiled across the FOXP2 promoter (R1 / R2
overlapping by ~15 bp at the anterior site, R3 ~210 bp upstream), profiled
with dCas9-APEX2 proximity proteomics in HEK293 cells.

The point of the walkthrough is to show the end-to-end pipeline:
load inputs → run the spatial deconvolution → inspect the binding-event
table → render the per-TF deconvolution deck. Heavy chunks
(`run_caspex()`, `run_caspex_extras()`) are not evaluated when this
vignette is built (they hit Ensembl, JASPAR and ChIP-Atlas over the
network and would slow `R CMD check` to a crawl), but every line is a
copy-paste-able recipe you can run interactively.

# Setup

```{r setup}
library(GLproxScape)
```

# 1. Load bundled inputs

GLproxScape ships the Mackenzie FOXP2 inputs under `inst/extdata/` so
this walkthrough is fully self-contained.

```{r inputs}
inputs_dir <- system.file("extdata/examples/foxp2_mackenzie",
                          package = "GLproxScape")
list.files(inputs_dir)
```

The input bundle has the canonical layout: a `grnas.tsv` manifest mapping
each region to its sgRNA protospacer and to the per-region proteomics
table, plus three `Region*.txt` files holding logFC + p-value columns.

```{r load-inputs}
inputs <- load_caspex_inputs(inputs_dir)
str(inputs, max.level = 1)
inputs$grnas
```

# 2. Anchor on the right transcript

FOXP2 has many alternative promoters; the canonical Ensembl FOXP2-201
(`ENST00000350908`) sits ~328 kb downstream of the cut sites Mackenzie
targeted, so we anchor on the alternative transcript whose TSS sits
right under their guide layout (`ENST00000901759`,
chr7:114,086,317; +1185/+1188/+973 from R1/R2/R3).

# 3. Run the spatial deconvolution

```{r run-caspex, eval = FALSE}
res <- run_caspex(
  gene             = "FOXP2",
  transcript       = "ENST00000901759",
  grnas            = inputs$grnas,
  data_files       = inputs$data_files,
  upstream         = 200,
  downstream       = 2000,
  out_dir          = tempfile("foxp2_run_"),
  weight_mode      = "lfc_signed",
  pval_thresh      = 0.5,
  motif_thresh     = 0.75,
  chipatlas        = FALSE   # set TRUE for the ChIP-Atlas overlay (~50 SRX/TF)
)
```

# 4. Inspect the result

The list returned by `run_caspex()` exposes (among other things)
`binding_events` (one row per called event), `spatial_df` (one row per
spatially-modelled TF), `pos_map` (gRNA TSS-relative positions),
`promoter_info` (the sequence and TSS offset), and `motif_results` (per-TF
JASPAR scan).

```{r inspect, eval = FALSE}
# Top motif-anchored events, sorted by amplitude
head(res$binding_events[order(-res$binding_events$weight), ], 10)

# How many events per TF?
table(res$binding_events$tf)

# A quick look at where MEIS1 was called
res$binding_events[res$binding_events$tf == "MEIS1",
                   c("position", "weight", "n_motifs_merged", "motif_based")]
```

# 5. Diagnostic deck

The full plot pack (motif-vs-NNLS scatter, TF co-occurrence triangle,
sigma sensitivity, jackknife stability, etc.) is one call:

```{r extras, eval = FALSE}
extras <- run_caspex_extras(
  res,
  out_dir = file.path(res$out_dir, "extras")
)
```

# 6. Switching anchors / weighting / network overlays

Two knobs that are worth knowing about for the FOXP2 reanalysis:

`motif_score_weight` lets you bias bubble amplitudes toward higher-PWM
motifs within each cluster. Default `"none"` is unweighted. `"linear"`
multiplies each candidate's weight by its score-fraction; `"log"` uses
`2^(score_frac - 1)`. Above any reasonable scan threshold the surviving
score range is narrow and predictions are robust to the choice — see the
`vignette("foxp2-mackenzie")` figures or just compare three runs
yourself by passing each value to `run_caspex()`.

`chipatlas = TRUE` adds an independent-validation lane below the
predicted-event bubbles. The first call downloads ~300 MB of metadata,
caches it under `tools::R_user_dir("caspex", "cache")`, and then fetches
per-SRX BED files on demand (small, also cached). Pass
`special_interest_gene = c("EP300", "CTCF", ...)` to bypass the per-TF
SRX cap for specific factors.

# Reproducing the paper figures

The paper's Figure 2 panels (MAX, FOXA1, GATA4/TFAP4, HMGB1) are pulled
from the per-TF binding-deconvolution deck (`binding_deconvolution.pdf`
in the run output directory). Figure S2A is the A5 TF co-occurrence
triangle from the extras pack. Figure S2C — the comparison across
`motif_score_weight` settings — is reproduced by running `run_caspex()`
three times with `motif_score_weight = "none" / "linear" / "log"` and
plotting the three event sets side-by-side.

# Citing GLproxScape

If you use GLproxScape in your work, please cite Ozcan et al. (in
preparation) and Mackenzie et al. 2026 for the underlying dataset.

# Session info

```{r session-info}
sessionInfo()
```
